---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)

library(tidyr)
library(ggplot2)
library(dplyr)
library(priorityqueue)
```

# priorityqueue

<!-- badges: start -->
![](https://img.shields.io/badge/cool-useless-green.svg)
[![R-CMD-check](https://github.com/coolbutuseless/priorityqueue/workflows/R-CMD-check/badge.svg)](https://github.com/coolbutuseless/priorityqueue/actions)
<!-- badges: end -->

`priorityqueue` is a package with two 
[priority queue](https://en.wikipedia.org/wiki/Priority_queue) implementations for R: 
one written in C, and the other written in plain R.  

Both of these are *max-priority-queues* i.e.
"popping" from the queue returns the value with highest priority.

These priority queues support any R object. The priority can be any finite `double()` value.


## What's in the box

Two priority queue implementations with identical call APIs.

* Pure R priority queue:
    * `rpq_create()`, `rpq_insert(rpq, obj, priority)`, `rpq_pop(rpq)`
* C-based priority queue:
    * `cpq_create()`, `cpq_insert(cpq, obj, priority)`, `cpq_pop(cpq)`


## Installation

You can install from [GitHub](https://github.com/coolbutuseless/priorityqueue) with:

``` r
# install.package('remotes')
remotes::install_github('coolbutuseless/priorityqueue')
```



## Example

Priority queues are a data-structure to store objects along with a priority with each object.

When *popping* a value off the queue, the object with the largest priority is
returned.

```{r example}
library(priorityqueue)

rpq <- rpq_create()
rpq_is_empty(rpq)

rpq_insert(rpq, obj = "#RStats", priority = 2)
rpq_insert(rpq, obj = "hello"  , priority = 3)
rpq_insert(rpq, obj = 2 * 2    , priority = 1)

rpq_length(rpq)

# Retrieve the objects from largest priority to smallest
rpq_pop(rpq)
rpq_pop(rpq)
rpq_pop(rpq)
```



## Rationale

I needed a data structure in R which would allow me to insert/pop items based upon
a *priority* value - what is known as a *priority queue*.

I wrote a quick solution in R that met my basic needs i.e. an environment 
containing a list of objects and a vector of values, plus some standard functions
to manipulate.  This code is the basis of the `rpq_*` functions in this package.

I then wondered how this could be done with C, and as a learning exercise I
wrote a minimal priority queue in C which supported R objects.  This was tidied
to be the C prioirty queue (`cpq_*`)  in this package.


## Timing

In general the C implementation is faster and uses less memory than the R 
implementation (no surprises there).

For larger queues, the C implementation can be ~10x faster and use 1/30th the memory.

This implementation of a priority queue compares favourably with the priority queue
included in the [{collections}](https://cran.r-project.org/package=collections) package
on CRAN.

<details>
<summary> Click to show/hide the timing code </summary>

```{r results='hide', warning=FALSE, message=FALSE}
r_priority_queue <- function(N) {
  set.seed(1)
  rpq <- rpq_create()
  
  for (i in 1:(N*2)) {
    rpq_insert(rpq, 1, runif(1))
    rpq_insert(rpq, 1, runif(1))
  }
  
  for (i in 1:N) {
    rpq_insert(rpq, 1, runif(1))
    rpq_insert(rpq, 1, runif(1))
    rpq_pop(rpq)
    rpq_pop(rpq)
    rpq_pop(rpq)
  }
}


cpq_priority_queue <- function(N) {
  set.seed(1)
  cpq <- cpq_create()
  
  for (i in 1:(N*2)) {
    cpq_insert(cpq, 1, runif(1))
    cpq_insert(cpq, 1, runif(1))
  }
  
  for (i in 1:N) {
    cpq_insert(cpq, 1, runif(1))
    cpq_insert(cpq, 1, runif(1))
    cpq_pop(cpq)
    cpq_pop(cpq)
    cpq_pop(cpq)
  }
}



collections_priority_queue <- function(N) {
  set.seed(1)
  colq <- collections::priority_queue()
  
  for (i in 1:(N*2)) {
    colq$push(1, runif(1))
    colq$push(1, runif(1))
  }
  
  for (i in 1:N) {
    colq$push(1, runif(1))
    colq$push(1, runif(1))
    colq$pop()
    colq$pop()
    colq$pop()
  }
}


res <- bench::press(
  N = c(10, 100, 1000),
  bench::mark(
    `Plain R priority queue` = r_priority_queue(N), 
    `C priority queue` = cpq_priority_queue(N),
    `collections::priority_queue()` = collections_priority_queue(N)
  )
)
```

</details>


```{r echo = FALSE, fig.height = 4}
summary(res) %>%
  select(expression, N, `itr/sec`, mem_alloc) %>%
  knitr::kable()

plot(res) + 
  theme_bw() + 
  theme(
    legend.position = 'bottom'
  )
```




## Technical Bits

Plain R priority queue

* Implemented as an environment with storage for a list of objects and a vector of priorities.
* New objects and their priorities are appended to the list/vector in the environment.
* When a value is *popped*, the objects are sorted by the priorities, and the
  first object is returned.
* The R version is slightly optimized for the situation where there are consecutive 
  reads, in that it does not re-sort the objects if no new objects have been
  added to the queue since the last pop operation.

C priority queue

* A heap-based priority queue.
* Creating a `cpq` creates an ExternalPointer to a `cpq_t` object which contains
  the queue and other information all entirely within C.
* An R list is attached to this ExternalPointer in C (as a *protected* item on the pointer).
  When new R objects are
  added to the priority queue, they are also added to this list to prevent
  them from being garbage collected.


## Related Software

* The [{collections}](https://cran.r-project.org/package=collections) package
  already includes a very good priority queue and bunch of other useful stuff.


## Acknowledgements

* R Core for developing and maintaining the language.
* CRAN maintainers, for patiently shepherding packages onto CRAN and maintaining
  the repository
